% latexmk -pdf week3

\documentclass[letterpaper]{article}

\input{setup}
\input{defns}

\title{15-791 ATPL \\ Week 3 Notes}
\author{Nathan Glover and Chase Norman}
\date{\today}
\begin{document}
\maketitle

\section{Exact Equality}

Previously, we defined the unary relation on terms $HT_A(M)$, also written as $M \in A$, denoting that $M$ satisfies the behavioral specification of hereditary termination given by $A$.
Along these lines, we will be defining a binary relation that can be thought of as $EQ_A(M, M')$, which we write $M \dot{=} M' \in A$, representing Exact (or Semantic) Equality in type $A$.
Again, we define this relation by induction on the structure of the closed type $A$. We redefine $M \in A$ to mean $M \dot{=} M \in A$. 
We are encouraged to view $A$ as a specification, making this construction an invocation of the ``propositions as types'' principle. 
Our ``theory of truth'' in defining the satisfaction of the proposition $M \dot{=} M' \in A$ will be based in the computation of $M$ and $M'$. 

\begin{definition}[Exact Equality]
    $$
  \begin{array}{cll}
      M \dot{=} M' \in \unitTy* & \Iff & M, M' \evalsTo \unitEx* \\
      M \dot{=} M' \in \ansTy* & \Iff & M, M' \evalsTo \yesEx* \text{ or } M, M' \evalsTo \noEx* \\
      M \dot{=} M' \in \prodTy*{A_1}{A_2} & \Iff &
         M \evalsTo \pairEx*{M_1}{M_2} \text{ and } M' \evalsTo \pairEx*{M_1'}{M_2'}\\ & & \text{ where } M_1 \dot{=} M_1' \in A_1 \text{ and } M_2 \dot{=} M_2' \in A_2 \\
      M \dot{=} M' \in \arrTy*{A_1}{A_2} & \Iff & M \evalsTo \lamEx*{x}{M_2} \text{ and } M' \evalsTo \lamEx*{x}{M_2'} \text{ where } M_1 \dot{=} M_1' \in A_1 \\ & & 
         \text{ implies } \Sub{M_1}{x}{M_2} \dot{=} \Sub{M_1'}{x}{M_2'} \in A_2 \\
  \end{array}
  $$
\end{definition}

These definitions essentially derive from our treatment of types as specifications. For each type, we reduce $M$ and $M'$ to values, and close under head expansion for the subterms. 
Note that our definition of exact equaltiy in the arrow type does not require $M_1$ and $M_1'$ to be identical, only semantically equal. 
The impact of this will become apparent as more concepts are added.

Now, we extend to open terms:


\begin{definition}[$\Gamma \gg M \dot{=} M' \in A$]
    $\gamma \dot{=} \gamma' \in \Gamma$ implies $\hat{\gamma} M \dot{=} \hat{\gamma'} M' \in A$ 
\end{definition}

Where $\gamma \dot{=} \gamma' \in \Gamma$ is variable-by-variable equality of the terms in substitutions $\gamma$ and $\gamma'$. 
Shortly, we will define judgemental equality $\equiv$ and prove that $\Gamma \entails{M \equiv N : A}$ implies $\Gamma \gg M \dot{=} N \in A$.
In essence, this is showing that proof implies truth. First, we establish some lemmas about exact equality.

\begin{lemma}[Head Expansion]\label{lem:headexpansion}
    If $M \dot{=} M' \in A$ and $N \to_\beta M$ then $N \dot{=} M' \in A$. Likewise, if $N' \to_\beta M'$ then $M \dot{=} N' \in A$. 
\end{lemma}
\begin{proof}
By induction on $A$, direct from the definition of exact equality.
\end{proof}

\begin{lemma}[Partial Equivalence Relation -- P.E.R.]
    For all $A$, exact equality in $A$ is symmetric and transitive. 
\end{lemma}
\begin{proof}
We prove these properties simultaneously, and by induction on $A$. We will require symmetry to prove transitivity in the case of the arrow type. 

\begin{enumerate}
    \item Case $A = \unitTy*$: Direct from the symmetry and transitivity of logical and.
    \item Case $A = \ansTy*$: Symmetry follows from the symmetry of logical and. For transitivity we observe that $M \evalsTo \yesEx*$ and $M \evalsTo \noEx*$
    are mutually exclusive for any term $M$. Therefore, if $M \dot{=} M' \in \ansTy*$ and $M' \dot{=} M'' \in \ansTy*$, we know $M$ and $M''$ must evaluate to
    the same value as $M$ and $M''$, completing the proof.
    \item Case $A = \prodTy*{A_1}{A_2}$: 
    \begin{enumerate}
        \item Apply symmetry from the inductive hypothesis on $M_1 \dot{=} M_1' \in A_1$ and $M_2 \dot{=} M_2' \in A_2$. The result follows directly.
        \item Apply transitivity from the inductive hypothesis on $M_1 \dot{=} M_1' \in A_1$ with $M_1' \dot{=} M_1''$ and $M_2 \dot{=} M_2'$ with $M_2' \dot{=} M_2''$.
        The result follows directly.
    \end{enumerate}
    \item Case $A = \arrTy*{A_1}{A_2}$
    \begin{enumerate}
        \item We want to show $M' \dot{=} M \in A$, so assume $M_1' \dot{=} M_1 \in A_1$ and let $M \evalsTo \lamEx*{x}{M_2}$ and $M' \evalsTo \lamEx*{x}{M_2'}$. 
        By the inductive hypothesis, we have $M_1 \dot{=} M_1' \in A_1$. With this and the given $M \dot{=} M' \in A$,
        we have $\Sub{M_1}{x}{M_2} \dot{=} \Sub{M_1'}{x}{M_2'} \in A_2$. Applying the inductive hypothesis on this equality completes the proof. 
        \item Using the P.E.R. trick, discussed below.
    \end{enumerate}
\end{enumerate}

To complete the transitivity case for $A = \arrTy*{A_1}{A_2}$, suppose $M \dot{=} M' \dot{=} M'' \in A$ and we want to show $M \dot{=} M'' \in A$. 
We have $M \evalsTo \lamEx*{x}{M_2}$, $M' \evalsTo \lamEx*{x}{M_2'}$, and $M'' \evalsTo \lamEx*{x}{M_2''}$. Suppose we have inputs $M_1 \dot{=} M_1'' \in A_1$. 
We must show $\Sub{M_1}{x}{M_2} \dot{=} \Sub{M_1''}{x}{M_2''} \in A_2$.

The natural approach would be to input $M_1 \dot{=} M_1'' \in A_1$ into the equalities from $M \dot{=} M' \dot{=} M'' \in A$, obtaining 
$\Sub{M_1}{x}{M_2} \dot{=} \Sub{M_1''}{x}{M_2'} \in A_2$ and $\Sub{M_1}{x}{M_2'} \dot{=} \Sub{M_1''}{x}{M_2''} \in A_2$. However, when we go to apply
transitivity on these, we find that the terms $\Sub{M_1''}{x}{M_2'}$ and $\Sub{M_1}{x}{M_2'}$ do not quite match. 

To remedy this, we require what we dub the ``P.E.R. trick.'' By symmetry, $M_1 \dot{=} M_1'' \in A_1$ implies $M_1'' \dot{=} M_1 \in A_1$.
By transitivity, we obtain that $M_1 = M_1 \in A_1$. This is the trick; any element that is related to another in a P.E.R. is also related to itself.

We then can apply $M_1 = M_1 \in A_1$ on $M \dot{=} M' \in A$ and $M_1 \dot{=} M_1'' \in A_1$ on $M' \dot{=} M'' \in A$ to obtain
$\Sub{M_1}{x}{M_2} \dot{=} \Sub{M_1}{x}{M_2'} \in A_2$ and $\Sub{M_1}{x}{M_2'} \dot{=} \Sub{M_1''}{x}{M_2''} \in A_2$. The two sides match now, 
and transitivity from the inductive hypothesis completes the proof.
\footnote{Analogously, we could have used the P.E.R. trick to find $M_1'' \dot{=} M_1'' \in A_1$ and applied it to $M' \dot{=} M'' \in A$.}
The P.E.R. trick allows us to bridge the gap between these two equalities, placing the transition from $M_1$ to $M_1''$ entirely on one side.  

\end{proof}

Reflexivity will hold if and only if a term is well-typed.

\begin{theorem}[Reflexivity]
    If $\Gamma \entails{\isOfTp{M}{A}}$ then $\Gamma \gg M \in A$, meaning $\Gamma \gg M \dot{=} M \in A$.
\end{theorem}
The proof of this was left as an exercise. It is necessary for this theorem to hold, otherwise the $R$ inference rule for judgemental equality is unsound.

\section{Judgemental Equality}\label{sec:jeq}

We define $\Gamma \entails{\isOfTp{M \equiv N}{A}}$ via inference rules. With the first set of rules, we axiomatically prescribe judgemental equality to be a P.E.R. with reflexivity
for all well-typed terms. 

\begin{mathpar}

\defrule[R][jeq:r]
{\Gamma \entails{\isOfTp{M}{A}}}
{ \Gamma \entails{\isOfTp{M \equiv M}{A}} }

\defrule[S][jeq:s]
{\Gamma \entails{\isOfTp{M \equiv N}{A}} }
{ \Gamma \entails{\isOfTp{N \equiv M}{A}} }

\defrule[T][jeq:t]
{\Gamma \entails{\isOfTp{M \equiv N}{A}} \\ \Gamma \entails{\isOfTp{N \equiv P}{A}} }
{ \Gamma \entails{\isOfTp{M \equiv P}{A}} }

\end{mathpar}

The second set of inference rules describe how to construct equivalent terms from equivalent subterms.

\begin{mathpar}

\defrule[Pair][jeq:pair]
{\Gamma \entails{\isOfTp{M_1 \equiv N_1}{A_1}} \\ \Gamma \entails{\isOfTp{M_2 \equiv N_2}{A_2}}}
{ \Gamma \entails{\isOfTp{\pairEx*{M_1}{M_2} \equiv \pairEx*{N_1}{N_2}}{\prodTy*{A_1}{A_2}}} }

\defrule[Proj][jeq:proj]
{\Gamma \entails{\isOfTp{M \equiv N}{\prodTy*{A_1}{A_2}}} }
{ \Gamma \entails{\isOfTp{M \cdot i \equiv N \cdot i}{A_i}}}

\defrule[Lam][jeq:lam]
{\Gamma, \isOfTp{x}{A_1} \entails{\isOfTp{M \equiv N}{A_2}} }
{ \Gamma \entails{\isOfTp{\lamEx*{x}{M} \equiv \lamEx*{x}{N}}{\arrTy*{A_1}{A_2}}}}

\defrule[App][jeq:app]
{\Gamma \entails{\isOfTp{M \equiv N}{\arrTy*{A_1}{A_2}}} \\ \Gamma \entails{\isOfTp{M_1 \equiv N_1}{A_1}} }
{ \Gamma \entails{\isOfTp{\appEx*{M}{M_1} \equiv \appEx*{N}{N_1}}{A_2}}}

\end{mathpar}

And finally, the third set of rules concern $\beta$ and $\eta$ equality.

\begin{mathpar}

\defrule[$\eta_1$][jeq:eta1]
{\Gamma \entails{\isOfTp{M}{\unitTy*}}}
{\Gamma \entails{\isOfTp{M \equiv \unitEx*}{\unitTy*}}}

\defrule[$\beta_\times$][jeq:betapair]
{\Gamma \entails{\isOfTp{M_1}{A_1}} \\ \Gamma \entails{\isOfTp{M_2}{A_2}}}
{\Gamma \entails{\isOfTp{\projEx*<i>{\pairEx*{M_1}{M_2}} \equiv M_i}{A_i}}}

\defrule[$\eta_\times$][jeq:etapair]
{\Gamma \entails{\isOfTp{M}{\prodTy*{A_1}{A_2}}}}
{\Gamma \entails{\isOfTp{\pairEx*{\projEx*<1>{M}}{\projEx*<2>{M}} \equiv M}{\prodTy*{A_1}{A_2}}}}

\defrule[$\beta_\to$][jeq:betaarr]
{\Gamma, \isOfTp{x}{A_1} \entails{\isOfTp{M_2}{A_2}} \\ \Gamma \entails{\isOfTp{M_1}{A_1}}}
{\Gamma \entails{\isOfTp{\appEx*{\lamEx*{x}{M_2}}{M_1} \equiv \Sub{M_1}{x}{M_2}}{A_2}}}

\defrule[$\eta_\to$][jeq:etaarr]
{\Gamma \entails{\isOfTp{M}{\arrTy*{A_1}{A_2}}}}
{\Gamma \entails{\isOfTp{\lamEx*{x}{\appEx*{M}{x}} \equiv M}{\arrTy*{A_1}{A_2}}}}

\end{mathpar}

We can now state the fundemental theorem for this logical relation.

\begin{theorem}[FTLR]
If $\Gamma \entails{\isOfTp{M \equiv N}{A}}$ then $\Gamma \gg M \dot{=} N \in A$.
\end{theorem}
\begin{proof}
We go by induction on the derivation of $\Gamma \entails{\isOfTp{M \equiv N}{A}}$. 
\begin{enumerate}
  \item For the first set of rules ($R$, $S$, and $T$) we make use of the reflexivity theorem as well as the lemma stating that exact equality is a P.E.R.
  \item For the second set of rules (\textsc{Pair}, \textsc{Proj}, \textsc{Lam}, \textsc{App}), the claim follows fairly easily from the definition of exact quality after applying the induction hypothesis. Occasionally, head expansion is necessary.
  \item For the third set of rules ($\eta_1$, $\beta_\times$, $\eta_\times$, $\beta_\to$, $\eta_\to$), we make use of the head expansion lemma and the reflexivity theorem.
\end{enumerate}
\end{proof}

\begin{remark}[Zig-Zag Closure]

Closed exact equality is zig-zag complete. That is, if $M \dot{=} M' \in A$, $N \dot{=} M' \in A$, and $N \dot{=} N' \in A$, then $M \dot{=} N' \in A$. This can be used instead of the P.E.R. trick to prove the FTLR. It will come up later in the course as a useful generalization of the P.E.R. trick that can handle settings where the types of the terms being judged equal are different.

\end{remark}


\section{More Types}

In this section, we'll extend the notion of exact equality to more types, namely $0$, $+$, $\natTy*$, and $\conatTy*$.

\begin{definition}[Exact Equality, Continued]
    $$
  \begin{array}{cll}
      M \dot{=} M' \in 0 & \Iff & \text{never} \\
      \\
      M \dot{=} M' \in \sumTy*{A_1}{A_2} & \Iff & \text{for some } i \in \{1, 2\}, M \evalsTo \injEx*<i>{M_i} \text{ and } M' \evalsTo \injEx*<i>{M'_i} \\
      & & \text{where $M_i \dot{=} M'_i$} \\
      \\
      M \dot{=} M' \in \natTy* & \Iff & \text{$\mathcal P(M, M')$, where $\mathcal P$ is the strongest relation} \\
      & & \text{such that $\mathcal P(M, M')$ if either $M, M' \evalsTo \zeroEx*$} \\
      & & \text{or $M \evalsTo \succEx*{N}$, $M' \evalsTo \succEx*{N'}$ with $\mathcal P(N, N')$} \\
      \\
      M \dot{=} M' \in \conatTy* & \Iff & \text{$\mathcal Q(M, M')$, where $\mathcal Q$ is the weakest relation} \\
      & & \text{such that $\mathcal Q(M, M')$ if either $\outEx*{M}, \outEx*{M'} \evalsTo \injEx*<1>{\unitEx*}$} \\
      & & \text{or $\outEx*{M} \evalsTo \injEx*<2>{N}$, $\outEx*{M'} \evalsTo \injEx*<2>{N'}$ with $\mathcal Q(N, N')$} \\
  \end{array}
  $$
\end{definition}

\end{document}